---
title: "Long assigment"
subtitle: "Analysis on e-commerce data"
author: "Pedro Miguel Sousa Magalhães"
date: November 2021
output: 
 html_notebook:
    toc: yes
    number_sections: yes
    toc_float: yes
    theme: cosmo
    highlight: tango
abstract: "A simple summary with conclusions"
fontfamily: arial
fontsize: 8pt
---

```{r, setup, include = FALSE } 
  
knitr::opts_knit$set(root.dir = "/projecto_intro_datascience")

```

```{r, libraries}

library(tidyverse)

```



```{r, importing data}

inbalanced_data <- read_csv("data/train_full.csv")

```


# Business understanding

*The company sells luxury items through its website and wants to improve success in its sales and better understand the behavior of the customers/users.*

Goals >> 1. what is convertion probability, 2. what variables influence conversion  

Torna-se um problema de classificação. Análise de pesos pode ser dada por modelo logistico ou por shapley

## Determine Business Objectives


## Assess situation


## Determine goals


# Data understanding

## Collect initial data


## Describe data


## Explore data


## Verify data quality


# Data preparation



# Modeling



# Evaluation and main conclusions




