---
title: "List & explain Statistical Measures of a distribution"
author: "Pedro Magalhães - 200202298 "
subtitle: Introduction to Data Science homework - 17/11/2021
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

```

# Population and sample 

All statistical measures present on this report will be calculated using a sample of 100 observations of a population of a continuous variable which closely follows a Beta distribution  with $\alpha = 1$ and $\beta = 5$.


```{r}
# set seed for random variables
set.seed(321)

# libraries
library(ggplot2)

# plot density function
pop <- data.frame(observation =rbeta(10000,1,5) * 80)

ggplot(pop, aes( x = observation )) + 
  geom_density(color = "red") +
  labs(title = "Population density plot") +
  theme_classic()

```

```{r}
set.seed(321)
sample <- data.frame(obs = round(rbeta(100,1,5) * 80, 0))

# ploting histogram for sample and density plot

sample_plot <- ggplot(sample, aes(x = obs )) +
  geom_histogram( bins = 10, colour = "black", fill = "white" ) +
  theme_classic() 

sample_plot + labs(title = "Sample histogram")

```

The data used on this exercise is therefore right skewed (eg: age distribution of a day center with kindergarden and elderly support) which affects the main calculation and interpretation of the main statistical measures.

# Measures of centrality

```{r include=FALSE}
# calculates the centrality measures for our sample data
measures <- summary(round(rbeta(100,1,5) * 80, 0))

mean <- as.numeric(measures["Mean"])
median <- as.numeric(measures["Median"])

# base R has no function for mode

Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

mode <- Mode(round(rbeta(100,1,5) * 80, 0))

```


## Mean

**Wikipedia definition:** In mathematics and statistics, the arithmetic mean, or simply the mean or the average (when the context is clear), is the sum of a collection of numbers divided by the count of numbers in the collection.

Therefore, given a sample of $\{x_1, x_2, ..., x_n \}$ the the average $\overline{x}$ is defined as follows

$$\overline{x} = \frac{1}{n}\sum_{i=1}^nx_i $$


```{r}
sample_plot +
  labs(title =  "Sample histogram with mean" ) +
  geom_vline(aes(xintercept = mean), color = "blue", linetype = "dashed", size = 1) +
  geom_text(aes(x = mean +1  ), label="mean", y = 30, angle = 90, colour = "blue")

```
Given the skewed nature of our sample data the interpretation of the mean should be done with caution since its not a suitable measure to describe the above distribution and does not provide a "true" center.

## Median

**Wikipedia definition:** In statistics and probability theory, the median is the value separating the higher half from the lower half of a data sample, a population, or a probability distribution. For a data set, it may be thought of as "the middle" value. 

The median represents the percentile 50 of a distribution


```{r}

sample_plot +
  labs(title =  "Sample histogram with median & mean" ) +
  geom_vline(aes(xintercept = mean), color = "blue", linetype = "dashed", size = 1) +
  geom_text(aes(x = mean + 1 ), label=paste("mean = ", mean) , y = 30, angle = 90, colour = "blue") +
  geom_vline(aes(xintercept = median), color = "green", linetype = "dashed", size = 1) +
  geom_text(aes(x = median + 1 ), label = paste("median = ", median) , y = 30, angle = 90, colour = "green")

```

## Mode

**Wikipedia definition:** The mode is the value that appears most often in a set of data values. 

In our scenario it shows us that most observations are of the value 5.

```{r}

sample_plot +
  labs(title =  "Sample histogram with median & mean" ) +
  geom_vline(aes(xintercept = mean), color = "blue", linetype = "dashed", size = 1) +
  geom_text(aes(x = mean + 1 ), label=paste("mean = ", mean) , y = 30, angle = 90, colour = "blue") +
  geom_vline(aes(xintercept = median), color = "green", linetype = "dashed", size = 1) +
  geom_text(aes(x = median + 1 ), label = paste("median = ", median) , y = 30, angle = 90, colour = "green") +
  geom_vline(aes(xintercept = mode), color = "red", linetype = "dashed", size = 1) +
  geom_text(aes(x = mode + 1 ), label = paste("mode = ", mode) , y = 30, angle = 90, colour = "red")

```


# Measures of dispersion

```{r include=FALSE}
min <- as.numeric(measures["Min."])
max <- as.numeric(measures["Max."])
range <- max - min
Q1 <- as.numeric(measures["1st Qu."])
Q3 <- as.numeric(measures["3rd Qu."])
IQR <- Q3 - Q1
sd <- sd(round(rbeta(100,1,5) * 80, 0))
var <- sd ^2

```


## Range

**Wikipedia definition:** In statistics, the range of a set of data is the difference between the largest and smallest values.[1] Difference here is specific, the range of a set of data is the result of subtracting the sample maximum and minimum.

```{r}
sample_plot +
  labs(title =  paste("Sample histogram with range of", range) ) +
  annotate("rect",xmin = min, xmax = max, ymin = 0, ymax = Inf, fill = "red", alpha = 0.2)
```


## Variance & Standard deviation

**Wikipedia definition:** In statistics, the standard deviation is a measure of the amount of variation or dispersion of a set of values.[1] A low standard deviation indicates that the values tend to be close to the mean (also called the expected value) of the set, while a high standard deviation indicates that the values are spread out over a wider range.

Therefore, given a sample of $\{x_1, x_2, ..., x_n \}$ the the standard deviation $\overline{x}$ is defined the squared root of the variation:

$$s = \sqrt{\frac{1}{n}\sum_{i=1}^n (x_i - \overline{x} )^2}  $$
Given the the skewed nature of our sample the varianace and standard deviation does not provide a "true" image on the dispersion around the mean. In order to understand the dispersion of this distibution it is advisable to use the Quartiles and inter quartiles information.


## Quartiles and InterQuartile Range

**Wikipedia definition:**  In statistics, a quartile is a type of quantile which divides the number of data points into four parts, or quarters, of more-or-less equal size. The data must be ordered from smallest to largest to compute quartiles; as such, quartiles are a form of order statistic.

```{r}
sample_plot +
  labs(title =  paste("Sample histogram with IQR of", IQR) ) +
  annotate("rect",xmin = Q1, xmax = Q3, ymin = 0, ymax = Inf, fill = "green", alpha = 0.2)
```
