---
title: "Teste intercalar - cadeira de Séries Temporais"
author: "Pedro Miguel Sousa Magalhães"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_notebook:
    theme: cosmo
    highlight: tango
  html_document:
    df_print: paged
    theme: cosmo
    highlight: tango
subtitle: Parte 2
---

```{r}
set.seed(123)

# import main libraries
library(astsa)
library(tidyverse)

```


## Generating random experience

```{r, run experience}
# initial index for numbered experience
n_exp <- data.frame( index = seq(1:1000) )

# data from theoratical model
data <- arima.sim(n = 100, model = list( c(1,0,0) , ar = 0.5) )

# random experience
experience <- function(index) {
  data <- arima.sim(n = 100, model = list( c(1,0,0) , ar = 0.5) ) # random AR process
  m <- arima(data, order = c(1,0,0)) # estimating parameters
  result <- m$coef[1] # extracting coeficients
  return(result)
}

# execute random experiences
coef_vector <- n_exp %>% pmap(experience) %>% bind_rows()

```


```{r fig.height=2}

tsplot(data, col = 4, ylab = "", main = expression(AR(1)~~~phi==0.5))

```

## Comparing with theoratical distribution

```{r}

mean = 0.5
sd = (1-0.5^2)/100

t <- rnorm(1000, mean, sd )

par(mfrow = c(2,2))

hist(coef_vector$ar1, main = "Random experience")
hist(t, main = "Theoretical distribution")
boxplot(coef_vector$ar1)
boxplot(t) 

```
```{r}
# summary statistics

summary(t)

summary(coef_vector$ar1)

```


## Commenting on the results

Based on the plots from both the random experience and a sample of 1000 observations from a theoretical distribution they both follow a normal distribution with mean around 0,48 - 0,5. Nonetheless, the random experience results shows a higher variance. Using  t-test for the difference of means with $HO: \mu_a - \mu_b = 0$ and a $\alpha = 0.05$ we reject the null hypothesis.

```{r}

t.test(coef_vector$ar1, t,  conf.level  = 0.95)

```
